---
title: "Final Project Experimenting"
subtitle: "Preliminary Exploratory Data Analysis"
author: "Jesse Quijano"
date: "2024-11-10"
output: 
  html_notebook:
    fig_height: 7
    fig_width: 10
---
# What Continent has the most meteorite strikes by quantity and volume?

### Set Up

#### Clean up environment and load packages:
```{r}
rm(list = ls())
library(tidyverse)

#install.packages("maps")
```
## The Data Sets
The primary data set I will use is from The National Aeronautics and Space Administration (NASA).

Each case, 45,716 in total, represents a meteorite that has "landed" on Earth. There are three primary variables I am interested in (at least for now):

* reclat - the latitude of where the meteorite was found.
* reclong - the longitude of where the metorite was found.
* mass - the weight in grams of the meteorite

The other data set is from Dtechtive. It's the first I found, but is now just serving as my back-up data. I'd like to find a way to link that data between this and the NASA data, but haven't found a reliable join as of yet.

```{r}
data <- read.csv("Meteorite_Landings_20241109.csv")
head(data)

rejects <- data %>% filter(is.na(reclong))

#install.packages("foreign")
library(foreign)
dbf <- read.dbf("cartodb_query.dbf") #another meteorite data set that might be useful as it contains links out to meteoritical society
head(dbf)
```

!! More information on the NASA data can be found at [NASA: Meteorite Landings](https://data.nasa.gov/Space-Science/Meteorite-Landings/gh4g-9sfh/about_data)

## Experiment Using `rworldmap` to Find Continent Names
The other data set I will definitely be using. Leveraging geo coordinates (lat/long) from a sample meteorite landing, see if I can find the corresponding country data - specifically the continent, but there seems to be some other interesting data I might be able to make use of.

```{r}
# will use this to convert the lat/long to continent
# Information found here: https://stackoverflow.com/a/14342127/1732853

library(sp)
#install.packages("rworldmap")
library(rworldmap)
vignette("rworldmap")
help("rworldmap")

countriesSP <- getMap(resolution='low')
head(countriesSP)

test_points <- data.frame(longitude = c(6.08333), latitude = c(50.77500))

pointsSP = SpatialPoints(test_points, proj4string=CRS(proj4string(countriesSP)))
indices = over(pointsSP, countriesSP)
indices
indices$REGION # will for sure need this!

coords_to_continent = function(points)
{
  print(points)
  countriesSP <- getMap(resolution='low')

  pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))
  indices = over(pointsSP, countriesSP)
  indices$REGION # will for sure need this!
}

coords_to_continent(test_points)

```

## Let's see if we can plot the data on the world map
```{r}
glyph_ready <- data %>%
  filter(!is.na(reclong), reclong <= 180, reclong >= -180) %>%
  mutate(continent = coords_to_continent(data.frame(longitude = c(reclong), latitude = c(reclat)))) %>% 
  select(lat = reclat, long = reclong, continent, weight = mass..g.)

world <- map_data("world")

ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region), color = "white", fill = "lightgray", linewidth = 0.1
  ) +
  geom_point(
    data = glyph_ready,
    aes(long, lat), #, color = by type maybe?
    alpha = 0.7,
    size = .1
  ) 
```

### Intuition Related Observations:

* Look into the 7,315 rows that are removed. They must be important?
* The observations in the U.S. must be better/more complete. The density of the strikes seems unusual.
* I think my scale might be off (which might explain the first bullet) - since I have a lot of points located in the ocean at the bottom of my plot.

## Fun with population
https://r2country.obi.obianom.com/

## More data experimentation
```{r}
data %>% group_by(recclass) %>% summarize(total = n()) %>% arrange(desc(total))
data %>% group_by(year) %>% summarize(total = n()) %>% arrange(desc(year))
```

